environmental_data_preperation

### 2018-10-05 src/biological_data_prep.Rmd
env. layer correction and dataset rebuilding complete

Load the env. data
```{r}
library(raster)
cell_layer <- raster("../output/env/glo_unique_cell.tif") #this is loading the cell layer that was created and saved as a tif
plot(cell_layer)
```
and just grab the attributes of the cell_layer
```{r}
cell_layer
```

@@

first make sure that the data_aea$depth without values == NA (currently they are marked xx)
```{r}
data_aea$depth[data_aea$depth == "xx"] <- NA #make all xx values NA
```

the depth data in data_aea is a factor. This needs to be numeric
```{r}
data_aea$depth <- as.numeric(as.character(data_aea$depth))
str(data_aea)
```

@@

>>>>

Load the env. data - you can just use the glorys layer or the biomer layer as they are identical
```{r}
library(raster)
cell_layer <- raster("../output/env/glo_unique_cell.tif") #this is loading the cell layer that was created and saved as a tif
plot(cell_layer)
```
and just grab the attributes of the cell_layer
```{r}
cell_layer
```

@@

the depth data in data_aea is a factor. This needs to be numeric
```{r}
data_aea$depth <- as.integer(as.character(data_aea$depth))
str(data_aea)
```

make sure that the data_aea$depth without values == NA (currently they are marked xx)
```{r}
data_aea$depth[data_aea$depth == "xx"] <- NA #make all xx values NA
```

@@

and save

```{r}
write.csv(data_aea,"../output/bio/data_aea_cell_amo_nao_bins.csv", row.names = FALSE)
```

### 2018-10-05 src/environmental_data_preperation.Rmd
env. layer correction and dataset rebuilding complete

# NetCDFs (Biomer and GLORYS)
read in a netcdf
note this is a GLORYS dataset
```{r}
library("ncdf4")
ncbr <- brick("../data/env/CMEMS_GLOBAL_REANALYSIS_PHY_001_025_salinity_monthly_mean_1993-01.nclonlat.ncdepthtrim.nc")
ncbr
```

reproject ...
There is currently an issue with the projectRaster command in the raster package. Michael Sumners kindly offered a [temporary solution](https://twitter.com/mdsumner/status/1040500830539935744)
```{r}
## use ncdf4 directly
library(ncdf4)
## connection to the file
nc <- nc_open("../data/env/CMEMS_GLOBAL_REANALYSIS_PHY_001_025_salinity_monthly_mean_1993-01.nclonlat.ncdepthtrim.nc")
## desired variable 
v <- ncvar_get(nc, "salinity")
## dimension coordinates
lon <- ncvar_get(nc, "longitude")
lat <- ncvar_get(nc, "latitude")
## a template, one map from the lon, lat, and v[,,1]
template <- raster(list(x = lon, y = lat, z = v[,,1]), crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0 ")

@@

Because the NA values are currently attributed with some other number and not NA in R (NA is set to -32767 in this particular layer), you need to make sure to exclude the cells which represent NA... so anything under 0.
the cell id will be a count of the cells with data. First calculate how many cells to remove from ncell (11,424)
```{r}
cell_val_freq <- freq(unique_cell, value=NULL)
cell_val_freq
```
ok so we have 4693 cells that have NA (or in reality a negative number that represents NA). so remove 4693 from ncell. We know that there are 11424 cells, soo... 11424-4693 = 6731 cells have data.

now give a unique ID to each cell
```{r}
unique_cell[unique_cell > 0] <- 1:6731 
writeRaster(unique_cell,'../output/env/unique_cell.tif',overwrite=TRUE, options=c('TFW=YES')) #so you can access it again
uniquecell_poly <- rasterToPolygons(aea[[1]], fun=NULL, n=4, na.rm=TRUE, digits=12, dissolve=FALSE)

writeOGR(uniquecell_poly, dsn = '.', layer = '../output/env/poly', driver = "ESRI Shapefile")

@@

ok so that unique cell layer was built from a GLORYS layer... in the biomer layer needs to match to this layer (and i have reason to think that it does not - the biomer layer was converted by me from an ORCA025 grid, and IIRC when I looked at the layers in ArcGIS there were some differences in the cell boundaries)

read in a biomer netcdf
note this only reads in the top layer which is what i need right now. To get the other layers, add ", lvar = 4" to brick (see code)
```{r}
library("ncdf4")
biobrick <- brick("../data/env/CMEMS_GLOBAL_REANALYSIS_BIO_001_018_chl_monthly_mean_1998-01.ncwgs.nc_lonlat.ncdepthtrim.nc") # lvar = 4 
biobrick
```

@@

Because the NA values are currently attributed with some other number and not NA in R (NA is set to -32767 in this particular layer), you need to make sure to exclude the cells which represent NA... so anything under 0.
the cell id will be a count of the cells with data. First calculate how many cells to remove from ncell (11,424) 
```{r}
bio_unique_cell <- bio_aea
bio_cell_val_freq <- freq(bio_unique_cell, value=NULL)
bio_cell_val_freq
```
ok so we have 4695 cells that have NA. Note that the GLORYS layer has 4693 NA cells....  a quick visual inspection with ArcGIS reveals one cell likes off the coast of the USA south of Nova Scotia (unique cell ID 4840) and one off the north tip of Nova Scotia (unique cell ID 3438)
so remove 4840 from ncell. We know that there are 11424 cells, soo... 11424-4695 = 6729 cells have data.

you need to remove cell 3438 and 4695 from the Glorys unique cell layer....
```{r}
values(unique_cell)[values(unique_cell) == 3438] = NA
values(unique_cell)[values(unique_cell) == 4840] = NA
```

look at the number of NAs again...
```{r}
uc_na <- freq(unique_cell,value=NA)
uc_na
```

ok great. ok so now give a unique ID to each BIOMER cell
```{r}
bio_unique_cell[bio_unique_cell > -1] <- 1:6729 
writeRaster(bio_unique_cell,'../output/env/bio_unique_cell.tif',overwrite=TRUE, options=c('TFW=YES')) #so you can access it again
bio_unique_cell
```
now give a unique ID to the GLORYS layer with the cells removed...
```{r}
glo_unique_cell <- unique_cell
glo_unique_cell[glo_unique_cell > -1] <- 1:6729 
writeRaster(glo_unique_cell,'../output/env/glo_unique_cell.tif',overwrite=TRUE, options=c('TFW=YES')) #so you can access it again
bio_unique_cell
```

and done (again)

>>>>

# NetCDFs (Biomer and GLORYS)
read in a netcdf
note this is a GLORYS dataset
```{r}
library("ncdf4")
ncbr <- brick("../data/env/1993_01_salinity.nc")
ncbr
```

reproject ...
There is currently an issue with the projectRaster command in the raster package. Michael Sumners kindly offered a [temporary solution](https://twitter.com/mdsumner/status/1040500830539935744)
```{r}
## use ncdf4 directly
library(ncdf4)
## connection to the file
nc <- nc_open("../data/env/1993_01_salinity.nc")
## desired variable 
v <- ncvar_get(nc, "salinity")
## dimension coordinates
lon <- ncvar_get(nc, "longitude")
lat <- ncvar_get(nc, "latitude")
## a template, one map from the lon, lat, and v[,,1]
template <- raster(list(x = lon, y = lat, z = v[,,1]), crs = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0 ")

@@

Because the NA values are currently attributed with some other number and not NA in R (NA is set to -32767 in this particular layer), you need to make sure to exclude the cells which represent NA... so anything under 0.
the cell id will be a count of the cells with data. First calculate how many cells to remove from ncell (15,402)
```{r}
cell_val_freq <- freq(unique_cell, value=NULL)
cell_val_freq
```
ok so we have 7634 cells that have NA (or in reality a negative number that represents NA). so remove 7634 from ncell. We know that there are 15402 cells, soo... 15402-7634 = 7768 cells have data.

now give a unique ID to each cell
```{r}
unique_cell[unique_cell > 0] <- 1:7768 
writeRaster(unique_cell,'../output/env/glo_unique_cell.tif',overwrite=TRUE, options=c('TFW=YES')) #so you can access it again
uniquecell_poly <- rasterToPolygons(aea[[1]], fun=NULL, n=4, na.rm=TRUE, digits=12, dissolve=FALSE)
writeOGR(uniquecell_poly, dsn = '.', layer = '../output/env/glo_unique_cell_poly', driver = "ESRI Shapefile")

@@

Read in a biomer netcdf
note this only reads in the top layer which is what i need right now. To get the other layers, add ", lvar = 4" to brick (see code)
```{r}
library("ncdf4")
biobrick <- brick("../data/env/1998_01_chl.nc") # lvar = 4 
biobrick
```

@@

Because the NA values are currently attributed with some other number and not NA in R (NA is set to -32767 in this particular layer), you need to make sure to exclude the cells which represent NA... so anything under 0.
the cell id will be a count of the cells with data. First calculate how many cells to remove from ncell (15402) 
```{r}
bio_unique_cell <- bio_aea
bio_cell_val_freq <- freq(bio_unique_cell, value=NULL)
bio_cell_val_freq
```
ok so we have 7652 cells that have NA. Note that the GLORYS layer has 7634 NA cells....  so let's remove the cells from the GLORYS raster that correspond to NA in the BIOMER raster...

```{r}
aea <- overlay(aea, bio_aea, fun = function(x, y){
  x[is.na(y[])] <- NA
  return(x)
})
plot(aea[[1]])
```

```{r}
aea_cut_cell_val_freq <- freq(aea[[1]], value=NULL)
aea_cut_cell_val_freq
```

great! So let's redo the GLORYS unique cell layer....

```{r}
unique_cell <- aea[[1]]
unique_cell[unique_cell > 0] <- 1:7750 
writeRaster(unique_cell,'../output/env/glo_unique_cell.tif',overwrite=TRUE, options=c('TFW=YES')) #so you can access it again
uniquecell_poly <- rasterToPolygons(aea[[1]], fun=NULL, n=4, na.rm=TRUE, digits=12, dissolve=FALSE)
writeOGR(uniquecell_poly, dsn = '.', layer = '../output/env/glo_unique_cell_poly', driver = "ESRI Shapefile")
unique_cell
```

and create a unique cell biomer layer..

```{r}
bio_unique_cell <- bio_aea
bio_unique_cell[bio_unique_cell > -1] <- 1:7750 
writeRaster(bio_unique_cell,'../output/env/bio_unique_cell.tif',overwrite=TRUE, options=c('TFW=YES')) #so you can access it again
bio_unique_cell
```
